The effects of uncertain context inference on motor adaptation

### 2022-01-12 adaptation/article/article.tex
Added Heald 2021. Modified for plos

\title{The effects of ongoing context inference on motor learning}
\shorttitle{Context-inference-dependent motor learning}
\author[1,2]{Cuevas Rivera, Darío}
\author[1,2]{Kiebel, Stefan J.}
\affil[1]{Chair of Neuroimaging, Faculty of Psychology, Technische Universität Dresden, 01062 Dresden, Germany.}
\affil[2]{Centre for Tactile Internet with Human-in-the-Loop (CeTI)}
\affiliation{~}

@@

\abstract{Abstract}
Humans have been shown to adapt their movements when a sudden change to the dynamics of the environment is introduced, a phenomenon called motor adaptation. If the change is reverted, the adaptation is also quickly reverted. Human are also able to adapt to multiple changes in dynamics presented separately, and to be able to switch between adapted movements on the fly. Such switching relies on contextual information which is often noisy or misleading, which affects the switch between adaptations. In this work, we introduce a computational model to explain the behavioral phenomena effected by imperfect contextual information. Specifically, we present a hierarchical model for motor adaptation based on exact Bayesian inference. This model explicitly takes into account contextual information and how the dynamics of context inference affect adaptation and action selection. We show how the proposed model provides a unifying explanation for four different experimentally established phenomena: (i) effects of sensory cues and proprioceptive information on switching between tasks, (ii) the effects of previously-learned adaptations on switching between tasks, in addition to (iii) the well-studied savings, de-adaptation and spontaneous recovery.

@@

In this work, we show that these unexplained phenomena can be explained in terms context inference, an idea that has recently been gaining traction in different fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}. We build on models such as MOSAIC \citep{Wolpert_Multiple_1998} and Bayesian switching models \citep{Kording_Bayesian_2004,Oh_Minimizing_2019} by expanding the concept of context and introducing an explicit mechanism for inferring the context based on Bayesian inference. In the proposed model, the context encompasses all the information relevant to making decisions, both environmental and internal to the agent making these decisions. To infer the current context, our model integrates several sources of information, including the predictions made by its internal forward models, an estimate of the reliability of these predictions, direct sensory information (such as visual contextual cues), proprioceptive information, and feedback provided by the environment.

@@

Our choice of an exact Bayesian learner greatly simplified our simulations, but we expect the results we presented here to hold using alternative models for motor adaptation, such as Kalman filter-based learners \cite[e.g.][]{Oh_Minimizing_2019,Baddeley_System_2003} and other Bayesian implementations \cite[e.g.][]{Wolpert_Multiple_1998,Kording_Bayesian_2004}. An exception is linear learners \cite[e.g.][]{Smith_Interacting_2006,Forano_Timescales_2020,Lee_Dual_2009}, as these models do not incorporate uncertainty in the estimate of the parameters of the forward models.

@@

The same rationale explains the results by \cite{Davidson_Scaling_2004}: motor learning, which in our model is modulated by context inference, is minimal for errors close to 2 and -2 (see their figure 2E). This is because an error of 2 or -2 signals that the participant incorrectly identified the context (as adaptation has a magnitude of 1).

>>>>

\title{The effects of uncertain context inference on motor adaptation}
\shorttitle{Context-inference-dependent motor adaptation}
\author[1,2]{Cuevas Rivera, Darío}
\author[1,2]{Kiebel, Stefan J.}
\affil[1]{Chair of Neuroimaging, Faculty of Psychology, Technische Universität Dresden, 01062 Dresden, Germany.}
\affil[2]{Centre for Tactile Internet with Human-in-the-Loop (CeTI)}
% \author[1]{Author list}
% \affil[1]{Affiliation list}
\affiliation{~}

@@

\abstract{Abstract}
Humans have been shown to adapt their movements when a sudden change to the dynamics of the environment is introduced, a phenomenon called motor adaptation. If the change is reverted, the adaptation is also quickly reverted. Human are also able to adapt to multiple changes in dynamics presented separately, and to be able to switch between adapted movements on the fly. Such switching relies on contextual information which is often noisy or misleading, which affects the switch between adaptations. In this work, we introduce a computational model to explain the behavioral phenomena effected by uncertain contextual information. Specifically, we present a hierarchical model for motor adaptation based on exact Bayesian inference. This model explicitly takes into account contextual information and how the dynamics of context inference affect adaptation and action selection. We show how the proposed model provides a unifying explanation for four different experimentally-established phenomena: (i) effects of sensory cues and proprioceptive information on switching between tasks, (ii) the effects of previously-learned adaptations on switching between tasks, (iii) the effects of training history on behavior in new contexts, in addition to (iv) the well-studied savings, de-adaptation and spontaneous recovery.

@@

In this work, we show that these unexplained phenomena can be explained in terms context inference, an idea that has recently been gaining traction in motor adaptation \citep{Heald_Contextual_2021}, as well as in other fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}. We build on models such as MOSAIC \citep{Wolpert_Multiple_1998} and Bayesian switching models \citep{Kording_Bayesian_2004,Oh_Minimizing_2019} by expanding the concept of context and introducing an explicit mechanism for inferring the context based on Bayesian inference. In the proposed model, the context encompasses all the information relevant to making decisions, both environmental and internal to the agent making these decisions. To infer the current context, our model integrates several sources of information, including the predictions made by its internal forward models, an estimate of the reliability of these predictions, direct sensory information (such as visual contextual cues), proprioceptive information, and feedback provided by the environment.

This work expands on the work by \cite{Heald_Contextual_2021}, who showed that context inference modulates learning rate in cases of unstable contexts. To do this, we focused on the effects of uncertain contextual information on switching behavior, especially during so-called error-clamp trials, in which errors are forced to zero by experimenters.

@@

Together with the results by \cite{Heald_Contextual_2021}, our results show that many behavioral phenomena observed across different experiments can be explained by the uncertainty in context inference and its effects on learning and action selection.

Our choice of an exact Bayesian learner greatly simplified our simulations, but we expect the results we presented here to hold using alternative models for motor adaptation, such as Kalman filter-based learners \cite[e.g.][]{Oh_Minimizing_2019,Baddeley_System_2003} and other Bayesian implementations \cite[e.g.][]{Wolpert_Multiple_1998,Kording_Bayesian_2004}, as well as the recently-introduced COIN model \citep{Heald_Contextual_2021}. An exception is linear learners \cite[e.g.][]{Smith_Interacting_2006,Forano_Timescales_2020,Lee_Dual_2009}, as these models do not incorporate uncertainty in the estimate of the parameters of the forward models.

@@

The same rationale explains the results by \cite{Herzfeld_memory_2014}: motor learning, which in our model is modulated by context inference, is minimal for errors close to 2 and -2 (see their figure 2E). This is because an error of 2 or -2 signals that the participant incorrectly identified the context (as adaptation has a magnitude of 1). Additionally, as was shown by \cite{Heald_Contextual_2021}, context inference explains the modulation of learning rate by the volatility of the environment observed by \cite{Herzfeld_memory_2014}.
